---
title: "Covid91 vaccine study Final2025"
author: "Aetan Barquero"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 5
date: "`r Sys.Date()`" 
---

```{r include=FALSE,echo=FALSE}
require(tidyverse)
require(tigerstats)
require(forcats)
FinalDat <-read.csv(file="finalSumr2024.csv",header=TRUE)
#recode to remove the annoying "X" from the dataframe
FinalData <- FinalDat  %>% select(-one_of("X"))
FinalData <- FinalData %>%
mutate(across(where(is.character), as.factor))
FinalData$treatment <- relevel(FinalData$treatment,ref="placebo")
#FinalData$infected <- relevel(FinalData$infected,ref="OK")
F = filter(FinalData, sex =="F")
M = filter(FinalData, sex =="M")
gay = filter(FinalData, LGBTQ =="gay")
Druguser = filter(FinalData, DrugUser =="yes")
```


Seperating the variables into four different categories (M=Male, F=Female, gay, and drug users) was imperent as this allows the data to be correctly analyzed. The seperation of these variables also allows understanding if one should take the Covid-19 vaccine. 

# Introduction

A new  preventative vaccine (covid921) is under test in clinical trials. Given the nature of this vaccine it is suspected that the drug may exhibit different effectiveness across the variety of subpopulations tracked. Consider population subgroups by gender( guys and gals), drug use, and sexual orientation. Write a complete report that separately considers this drug for each subgroup and your recommendations as to whether people in the subgroup should take the vaccine. Structure your report in four sections each devoted to one of the four subgroups, use null and alternative hypothesis statements and correct statistical verbage when referencing p-values in conclusions.

## Males


```{r}
dd2 <- M %>% group_by(infected,treatment) %>% summarize(count=n()) %>% mutate(prcnt=count/sum(count))


# the group_by followed by summarize(count=n())

basicC <- ggplot(dd2,aes(x=infected,y=count,fill=treatment))
basicC + geom_bar(stat="identity",position="dodge")

#Now for percentage plot

basicCC <- ggplot(dd2,aes(x=infected,y=prcnt*100,fill=treatment)) 
basicCC + geom_bar(stat="identity", position = "dodge")
```


The plot above displays the population of male people the if they caught and took the vaccine or had a placebo.

Out of all the males who caught Covid, the ones who took the placebo are approximately around 75% and the men who took the vaccine was exceedingly small around 25%. Of the people who did fine, it was split 50/50 between people who took the placebo vs. the people who had the vaccine. 

The barplots were generated using three main objects: dd2, basicC, and basicCC. The first object, dd2, appears in the Environment pane and contains four variables: treatment, infected, count, and prcnt (representing percent). This dataset is created using group_by() and summarize() to count the number of observations in each treatment–infection group, followed by mutate() to calculate the corresponding proportions.

The second object, basicC, also visible in the Environment pane, includes 11 variables and is used to create the first barplot that displays raw counts across groups. The final object, basicCC, located in the Files pane, also contains 11 variables and is used to generate the second barplot, which visualizes the percentages of infected individuals by treatment group.

### Numerical 

```{r}
table1 <- xtabs(~infected + treatment, data=M)
rowPerc(table1)
colPerc(table1)
```

These numerical results summarize the percentages of males who became infected with COVID-19 after receiving either the placebo or the vaccine. Among those who contracted COVID-19, 73.80% had taken the placebo, while only 26.20% had taken the vaccine, indicating a noticeable reduction in infections among vaccinated individuals. For the males who remained healthy, the percentages were nearly evenly split: 50.01% of the placebo group and 49.99% of the vaccine group were not infected. Overall, these results suggest that the vaccine is effective for males, as the proportion of infections is much lower among those who received the vaccine compared with the placebo group.

## Females


## LGBTQ



## Druggies


# Overall Results and Conclusions